The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-01 manuscript/manuscript.Rnw
include top dog and cow genes

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage[plain]{fancyref}
\renewcommand*{\fref}[1]{\Fref{#1}}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\pagestyle{fancy}

@@

% Acronyms
\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}
\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads \cite{Trapnell.ea2010:Transcriptassemblyandquantification}}

@@

\subsection{Transcriptome Assembly and Mapping}

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \fref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
(\fref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM} were quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \textit{Pan paniscus}. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

@@

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- combined.fpkm[species==species.name &
                      !grepl("^MT-?",gene_short_name)][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
@

@@

\begin{wraptable}{I}{0.3\textwidth}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Homo sapiens}}
\label{tab:top50_human_placenta}
\end{wraptable}

@@

\begin{wraptable}{O}{0.3\textwidth}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Pan paniscus}}
\label{tab:top50_bonobo_placenta}
\end{wraptable}

@@

@@

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage[plain]{fancyref}
\renewcommand*{\fref}[1]{\Fref{#1}}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

% Acronyms
\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}
\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads}

@@

\subsection{Transcriptome Assembly, Alignment, and Quantification}

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \fref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
(\fref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \textit{Pan paniscus}. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

@@

\begin{figure}
<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
print(ggplot(star.logs[grepl("%",field),],
             aes(x=species,y=as.numeric(gsub("%","",value)),
                 color=field))
      + geom_boxplot()
      + facet_grid(field~.,scales="free")
      + ylab("Percent of Reads")
      + theme(legend.position="top",
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1))
      )
@ 
\caption{Transcript statistics for each species}
\label{fig:transcriptome_statistics}
\end{figure}

@@

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- combined.fpkm[species==species.name &
                      !grepl("^MT-?",gene_short_name) & 
                          !grepl("(RNA\\d|MT)-",human_name)
                          ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
@

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Homo sapiens}}
\label{tab:top50_human_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.bonobo.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.bonobo.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Pan paniscus}}
\label{tab:top50_bonobo_placenta}
\end{wraptable}

@@

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.dog.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Canis familiaris}}
\label{tab:top50_dog_placenta}
\end{wraptable}

@@

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
top.cow.genes <- top.genes.table(species.name="bos taurus")
print(xtable(top.cow.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.cow.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Bos taurus}}
\label{tab:top50_cow_placenta}
\end{wraptable}
